---
title: "Impact of Late First-Half Goals on Match Outcomes"
author: ""
date: "2025-12-01"
output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies:
      fontspec: null
mainfont: Arial
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(nnet)
library(lubridate)
```

1.  Data Loading and Initial Inspection We begin by loading the raw
    dataset events.csv. This step includes error handling to ensure the
    file exists and inspecting the raw structure of the data (column
    types, missing values) to determine necessary cleaning steps.

```{r}
# Load data; file name is 'events.csv'
tryCatch({
  df <- readr::read_csv("events.csv")
  print("Data loaded successfully.")
}, error = function(e) {
  print("Error: Could not load the file.")
  df <- NULL
})

# Missing Value Check
na_counts <- colSums(is.na(df))
na_counts_df <- data.frame(
  Variable = names(na_counts),
  Missing_Count = na_counts,
  Missing_Percentage = round((na_counts / nrow(df)) * 100, 2)
)
print(head(na_counts_df[na_counts_df$Missing_Count > 0, ]))
```

2.  Data Cleaning and Type Conversion The raw data imports columns as
    characters or generic types. Here, we convert scores and periods to
    numeric types and parse the EventName to separate the competing
    teams. We also ensure the EventDate is properly formatted for any
    time-series analysis.

```{r}
df_clean <- df %>%
  mutate(
    Score1 = as.numeric(Score1),
    Score2 = as.numeric(Score2),
    Period1 = as.numeric(Period1), # Goals scored in the 1st half
    Period2 = as.numeric(Period2), # Goals scored in the 2nd half
    FirstGoalTeam = as.numeric(FirstGoalTeam),
    FirstGoalMinute = as.numeric(FirstGoalMinute)
  ) %>%
  separate(
    col = EventName,
    into = c("HomeTeam", "AwayTeam"),
    sep = " - ",
    remove = FALSE,
    extra = "merge" 
  ) %>%
  mutate(
    EventDate = lubridate::ymd_hms(EventDate)
  )

cat("Cleaned dataframe dimensions:", dim(df_clean))
```

3.  Feature Engineering: Identifying Late Goals We define a "Late
    First-Half Goal" as a goal scored between the 41st and 50th minute
    (inclusive of stoppage time).

```{r}
# Function to check for a late first-half goal (41st min up to 50th min)
check_late_goal <- function(goal_minutes_string) {
  if (is.na(goal_minutes_string) || goal_minutes_string == "NULL") {
    return(0)
  }

  # Split multiple goals by comma
  raw_tokens <- unlist(strsplit(goal_minutes_string, ","))
  
  # Helper to parse "45+2" into 47
  parse_minute <- function(x) {
    if (grepl("\\+", x)) {
      parts <- as.numeric(unlist(strsplit(x, "\\+")))
      return(sum(parts))
    } else {
      return(as.numeric(x))
    }
  }
  
  minutes <- suppressWarnings(sapply(raw_tokens, parse_minute))

  # Check range
  if (any(minutes >= 41 & minutes <= 50, na.rm = TRUE)) {
    return(1) 
  } else {
    return(0) 
  }
}

df_features <- df_clean %>%
  rowwise() %>%
  mutate(
    LateFH_Goal_Scored = check_late_goal(Team1GoalMinutes),
    LateFH_Goal_Conceded = check_late_goal(Team2GoalMinutes)
  ) %>%
  ungroup() %>%
  mutate(
    Match_Outcome = case_when(
      Score1 > Score2 ~ "Win",
      Score1 < Score2 ~ "Loss",
      Score1 == Score2 ~ "Draw",
      TRUE ~ NA_character_
    ),
    Match_Outcome = factor(Match_Outcome, levels = c("Draw", "Win", "Loss"))
  )
```

4.  Filtering for Top 5 Leagues To ensure consistency in competition
    level, we filter the dataset to include only matches from the "Top
    5" European leagues. We handle multi-language variations (English
    and Russian/Cyrillic) commonly found in data scrapers.

```{r}
# Filtering data
top_5_leagues_variants <- c(
    "Premier League", "Премьер-Лига",
    "Primera Division", "Примера", "Примера Дивизион",
    "Серия А", "Serie A",
    "Bundesliga", "Бундеслига",
    "Ligue 1", "Лига 1"
)

df_filtered <- df_features %>%
  filter(ChampionshipSubGroupName %in% top_5_leagues_variants) %>%
  mutate(
    Standard_League_Name = case_when(
      ChampionshipSubGroupName %in% c("Premier League", "Премьер-Лига") ~ "Premier League",
      ChampionshipSubGroupName %in% c("Primera Division", "Примера", "Примера Дивизион") ~ "La Liga",
      ChampionshipSubGroupName %in% c("Серия А", "Serie A") ~ "Serie A",
      ChampionshipSubGroupName %in% c("Bundesliga", "Бундеслига") ~ "Bundesliga",
      ChampionshipSubGroupName %in% c("Ligue 1", "Лига 1") ~ "Ligue 1",
      TRUE ~ "Other"
    ),
    Standard_League_Name = factor(Standard_League_Name)
  )

cat("Matches remaining after filtering:", nrow(df_filtered))
```

5.  Visual Analysis Before modeling, we visualize the data to understand
    the distribution of outcomes and the prevalence of late goals.

Distribution of Match Outcomes by League

```{r}
ggplot(df_filtered, aes(x = Standard_League_Name, fill = Match_Outcome)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Match Outcome Distribution by League",
    y = "Percentage",
    x = "League",
    fill = "Home Outcome"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

6.  Statistical Modeling (Multinomial Regression) We use Multinomial
    Logistic Regression to model the probability of a Win, Loss, or
    Draw.

Model Note: We use the Home Team perspective for modeling. We do not
duplicate the data (flipping to add Away rows) because this would
violate the statistical assumption of independence (a Home Win
guarantees an Away Loss in the same match).

Variables:

Late Goal Scored: Did the home team score in the 41-50 min window?

Late Goal Conceded: Did the away team score in that window?

Halftime Goals: Control variable for the current state of the match.

League: Control variable for league-specific trends.

```{r}
# We use the Home Team perspective (df_filtered) directly
# Rename cols for clarity in the model output
df_model <- df_filtered %>%
  mutate(
    Home_Late_Goal = LateFH_Goal_Scored,
    Away_Late_Goal = LateFH_Goal_Conceded,
    Home_Goals_HT = Period1, # Assuming Period1 is Home Goals
    Away_Goals_HT = Period2  # Assuming Period2 is Away Goals
  )

model <- multinom(
    Match_Outcome ~ 
        Home_Late_Goal + 
        Away_Late_Goal +
        Home_Goals_HT + 
        Away_Goals_HT + 
        Standard_League_Name,
    data = df_model
)

cat("\n### Model AIC ###\n")
print(model$AIC)
```

7.  Interpretation of Results We calculate Odds Ratios (OR) and P-values
    to interpret the results.

OR \> 1: Increases the likelihood of the outcome (relative to Draw).

OR \< 1: Decreases the likelihood.

P-value \< 0.05: Statistically significant.

```{r}
# Calculate Z-statistics and P-values
z_stats <- summary(model)$coefficients / summary(model)$standard.errors
p_values <- (1 - pnorm(abs(z_stats))) * 2

# Calculate Odds Ratios
odds_ratios <- exp(coef(model))

print("--- Odds Ratios (Likelihood vs. Draw) ---")
print(round(odds_ratios, 3))

print("--- P-Values ---")
print(round(p_values, 4))
```
